Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2025-01-07 paper-new/concrete-relational-model.tex
check in paper changes as of camera-ready submission on 11/22/24

We will generally write $A$ for both a predomain and its underlying set; if we
need to emphasize the difference we will write $|A|$ for the underlying set of
$A$.
%
Given a predomain $A$, we can form the predomain $\later A$. The
underlying set is $\later |A|$ and the ordering is given by $\tilde{x}
\ltdyn_{\later A} \tilde{x'}$ iff $\later_t(\tilde{x}_t \ltdyn_A
\tilde{x'}_t)$, and likewise for bisimilarity.

We similarly enhance our error domains with ordering and bisimilarity
relations, but we further need to impose conditions on the algebraic
structure: the error element should be the least element (so that it
models graduality), the think map should be monotone and bisimilarity-preserving,
and the bisimilarity relation should relate $\delta$ to the identity
(so that it models bisimilarity).
%
\begin{definition}
An \textbf{error domain} $B$ consists of a predomain $B$ along with the following data:
\begin{itemize}
    \item A distinguished ``error'' element $\mho_B \in B$ such that $\mho \ltdyn_B x$ for all $x$
    \item A morphism of predomains $\theta_B \colon \later B \to B$
    \item For all $x, y : B$ with $x \bisim_B y$, we have $\theta_B(\nxt\, x) \bisim_B y$
\end{itemize}
A \textbf{morphism of error domains} is a morphism of the underlying
predomains that preserves the error element and the $\theta$ map.
\end{definition}
For an error domain $B$, we define the predomain morphism $\delta_B := \theta_B
\circ \nxt$.

@@

The most interesting type to interpret is the dynamic
type. The beauty of guarded domain theory is that once we have define
$\laterhs$ for predomains, we can solve the same domain equation as in
Section~\ref{sec:concrete-term-model} changed only in that we interpret it as a
domain equation on predomains rather than sets:
%
\[ D \cong \mathbb{N}\, + (D \times D)\, + \laterhs U(D \arr \li D) \]
The complete definition of the ordering $\ltdyn_D$ is provided in the appendix
for completeness. We refer to the three injections as $e_\nat$, $e_\times$,
and $e_\to$, respectively.

@@

To accomplish this, we equip the predomains $A$ in our semantics with
a monoid $M_A$ of \emph{syntactic perturbations}, as well as a means
of \emph{interpreting} these syntactic perturbations as
\emph{semantic} perturbations on $A$, i.e., as endomorphisms that are
bisimilar to the identity. We call such an object a \emph{value
object}, and this will serve as the final definition of a denotation of
a type in our semantics. Similarly, we define computation objects to be
error domains equipped with a notion of syntactic perturbation:
\begin{definition}[value and computation objects + morphisms]
  A \textbf{value object} consists of a predomain $A$, a monoid $M_A$ and a
  homomorphism of monoids $i_A \colon M_A \to \morbisimid{A} := \{ f : A \to A \mid f
  \bisim \id_A \}$.

@@

Having defined syntactic perturbations and value and computation objects, we can
now make formal the final aspect of the model: the notion of
quasi-representability of relations.
%
\begin{definition}[quasi-left-representable relations]\label{def:quasi-left-representable}
Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects, and let $c : A
\rel A'$ be a predomain relation. We say that $c$ is
\emph{quasi-left-representable} by a predomain morphism $e_c : A \to A'$ if there
are perturbations $\delle_c \in M_A$ and $\delre_c \in M_{A'}$ such that the
following two squares exist: 
(1) $\upl$: $e_c \ltsq{c}{r(A')} i_{A'}(\delre_c)$, and
(2) $\upr$: $i_A(\delle_c) \ltsq{r(A)}{c} e_c$.
% If $c$ is quasi-left-representable by $e_c$, we call $e_c$ an \emph{embedding} morphism.

@@

Likewise, we define quasi-right-representability as follows:
%
\begin{definition}[quasi-right-representable relations]\label{def:quasi-right-representable}
Let $(B, M_B, i_B)$ and $(B', M_{B'}, i_{B'})$ be computation objects, and let
$d : B \rel B'$ be an error domain relation. We say that $d$ is
\emph{quasi-right-representable} by an error domain morphism $p_d : B' \multimap B$ if
there are perturbations $\dellp_d \in M_B$ and $\delrp_d \in M_{B'}$ such that
the following two squares exist: 
(1) $\dnl$: $p_d \ltsq{r(B')}{d} i_{B'}(\delrp_d)$, and
(2) $\dnr$: $i_B(\dellp_d) \ltsq{d}{r(B)} p_d$.
% If $d$ is quasi-right-representable by $p_d$, we call $p_d$ a \emph{projection} morphism.

@@

For ordinary representability, the relation is \emph{uniquely
determined} by any morphism that left- or right-represents it: if two
relations are left-represented by the same morphism, then they are
equal. But for quasi-representability, this is no longer the case due
to the presence of perturbations. Instead we have the following weaker property:
\begin{definition}[quasi-equivalence]\label{def:quasi-equivalent}
  Let $A$ and $A'$ be value objects, and let $c, c' : A \rel A'$ be relations
  between the underlying predomains. We say that $c$ and $c'$ are
  \textbf{quasi-equivalent}, written $c \qordeq c'$, if there exist
  syntactic perturbations $\delta^l_1, \delta^l_2 \in M_A$ and $\delta^r_1,
  \delta^r_2 \in M_{A'}$ such that there is a square $i_A(\delta^l_1)
  \ltsq{c}{c'} i_{A'}(\delta^r_1)$ and a square $i_A(\delta^l_2) \ltsq{c'}{c}
  i_{A'}(\delta^r_2)$. Given computation objects $B$ and $B$, we make the
  analogous definition for relations $d, d' : B \rel B'$ between the underlying
  error domains.
\end{definition}

@@

We can now give the final definition of value and computation relations for our
model, which serve as the denotation of our type precision derivations:
%
\begin{definition}[value relations]
A \textbf{value relation} between value objects $(A, M_A, i_A)$ and $(A',
M_{A'}, i_{A'})$ consists of a predomain relation $c : A \rel A'$ that has a
push-pull structure, is quasi-left-representable by a morphism $e_c : A \to A'$,
and is such that $\li c$ is quasi-right-representable by an error domain morphism
$p_c : \li A' \to \li A$.

@@

To model the term precision in a way that is oblivious to the stepping
behavior of terms, we need to weaken the notion of square to allow for the
morphisms on either side to be ``out of sync'' rather than in lock step.
In the below, $f' \bisimsq{A_i}{A_o} f$ is the natural extension of
bisimilarity to morphisms: given bisimilar inputs, the functions have
bisimilar outputs.
\begin{definition}
  Let $c_i : A_i \rel A_i'$, $c_o : A_o \rel A_o'$ $f : A_i \to A_o$
  and $g : A_i' \to A_o'$.  An extensional square $f
  \ltsqbisim{c_i}{c_o} g$ holds if there \emph{exist} $f'
  \bisimsq{A_i}{A_o} f$ and $g' \bisimsq{A_i'}{A_o'} g$ such that $f' \ltsq{c_i}{c_o} g'$.
\end{definition}

@@

The term semantics $\sem{M}$ is nearly identical to the term semantics in
Section~\ref{sec:concrete-term-model}. One difference is that the interpretation
of the function type $A_i \ra A_o$ in the term semantics of Section
\ref{sec:concrete-term-model} involves \emph{arbitrary functions}, while the
interpretation of the function type in the term semantics defined here involves
only those functions that are morphisms of predomains. Additionally, the
semantics of casts is slightly different. In the semantics of Section
\ref{sec:concrete-term-model}, the projection corresponding to the type
precision $cc'$ is simply the composition of the projections. In the new
semantics, the corresponding projection involves a perturbation.  
%
% EXTENDED VERSION
% This is a consequence of the construction of the composition of
% value/computation relations (the relevant projection is defined in the proof
% of Lemma \ref{lem:representation-comp-F-U} in the Appendix).
%
% CONFERENCE VERSION
This is a consequence of the construction of the composition of
value/computation relations (the relevant projection is defined in the appendix
of the extended version of this paper \cite{gtt-sgdt-extended-version}).
%
However, the two projections are extensionally equivalent since they are the
same up to a perturbation.

>>>>

We will generally write $A$ for both a predomain and its underlying set; if we
need to emphasize the difference we will write $|A|$ for the underlying set of
$A$.
%
Given a predomain $A$, we can form the predomain $\later A$. The
underlying set is $\later |A|$ and the ordering is given by $\tilde{x}
\ltdyn_{\later A} \tilde{x'}$ iff $\later_t(\tilde{x}_t \ltdyn_A
\tilde{x'}_t)$, and likewise for bisimilarity.
%
% Given predomains $A_i$ and $A_o$ and morphisms $f, g : A_i \to A_o$, we define
% $f \bisimsq{A_i}{A_o} g$ to be the natural extension of bisimilarity to
% morphisms: given bisimilar inputs, the morphisms have bisimilar outputs.

We similarly enhance our error domains with ordering and bisimilarity
relations, but we further need to impose conditions on the algebraic
structure: the error element should be the least element (so that it
models graduality), the think map should be monotone and bisimilarity-preserving,
and the bisimilarity relation should relate $\delta$ to the identity
(so that it models bisimilarity).
%
\begin{definition}
An \textbf{error domain} $B$ consists of a predomain $B$ along with the following data:
\begin{itemize}
    \item A distinguished ``error'' element $\mho_B \in B$ such that $\mho \ltdyn_B x$ for all $x$
    \item A morphism of predomains $\theta_B \colon \later B \to B$
    \item For all $x, y : B$ with $x \bisim_B y$, we have $\theta_B(\nxt\, x) \bisim_B y$
    % \item We have $\theta_B \circ \nxt \bisimsq{B}{B} \id$
\end{itemize}
A \textbf{morphism of error domains} is a morphism of the underlying
predomains that preserves the error element and the $\theta$ map.
\end{definition}
For an error domain $B$, we define the predomain morphism $\delta_B := \theta_B
\circ \nxt$.

@@

The most interesting type to interpret is the dynamic
type. The beauty of guarded domain theory is that once we have define
$\laterhs$ for predomains, we can solve the same domain equation as in
Section~\ref{sec:concrete-term-model} changed only in that we interpret it as a
domain equation on predomains rather than sets:
%
\[ D \cong \mathbb{N}\, + (D \times D)\, + \laterhs U(D \arr \li D) \]
%
% EXTENDED VERSION
% The complete definition of the ordering $\ltdyn_D$ is provided in the appendix
% for completeness.
%
% CONFERENCE VERSION
The complete definition of the ordering $\ltdyn_D$ is provided in the appendix
of the extended version of this paper \cite{gtt-sgdt-extended-version}.
%
We refer to the three injections as $e_\nat$, $e_\times$,
and $e_\to$, respectively.

@@

To accomplish this, we equip the predomains $A$ in our semantics with
a monoid $M_A$ of \emph{syntactic perturbations}, as well as a means
of \emph{interpreting} these syntactic perturbations as
\emph{semantic} perturbations on $A$, i.e., as endomorphisms that are
bisimilar to the identity. We call such an object a \emph{value
object}, and this will serve as the final definition of a denotation of
a type in our semantics. Similarly, we define computation objects to be
error domains equipped with a notion of syntactic perturbation:
\begin{definition}
  A \textbf{value object} consists of a predomain $A$, a monoid $M_A$ and a
  homomorphism of monoids $i_A \colon M_A \to \morbisimid{A} := \{ f : A \to A \mid f
  \bisim \id_A \}$.

@@

Having defined syntactic perturbations and value and computation objects, we can
now make formal the final aspect of the model: the notion of
quasi-representability of relations.
%
\begin{definition}\label{def:quasi-left-representable}
Let $(A, M_A, i_A)$ and $(A', M_{A'}, i_{A'})$ be value objects, and let $c : A
\rel A'$ be a predomain relation. We say that $c$ is
\textbf{quasi-left-representable} by a predomain morphism $e_c : A \to A'$ if there
are perturbations $\delle_c \in M_A$ and $\delre_c \in M_{A'}$ such that the
following two squares exist: 
(1) $\upl$: $e_c \ltsq{c}{r(A')} i_{A'}(\delre_c)$, and
(2) $\upr$: $i_A(\delle_c) \ltsq{r(A)}{c} e_c$.
% If $c$ is quasi-left-representable by $e_c$, we call $e_c$ an \emph{embedding} morphism.

@@

Likewise, we define quasi-right-representability as follows:
%
\begin{definition}\label{def:quasi-right-representable}
Let $(B, M_B, i_B)$ and $(B', M_{B'}, i_{B'})$ be computation objects, and let
$d : B \rel B'$ be an error domain relation. We say that $d$ is
\textbf{quasi-right-representable} by an error domain morphism $p_d : B' \multimap B$ if
there are perturbations $\dellp_d \in M_B$ and $\delrp_d \in M_{B'}$ such that
the following two squares exist: 
(1) $\dnl$: $p_d \ltsq{r(B')}{d} i_{B'}(\delrp_d)$, and
(2) $\dnr$: $i_B(\dellp_d) \ltsq{d}{r(B)} p_d$.
% If $d$ is quasi-right-representable by $p_d$, we call $p_d$ a \emph{projection} morphism.

@@

For ordinary representability, the relation is \emph{uniquely
determined} by any morphism that left- or right-represents it: if two
relations are left-represented by the same morphism, then they are
equal. But for quasi-representability, this is no longer the case due
to the presence of perturbations. Instead we have the following weaker property:
\begin{definition}\label{def:quasi-equivalent}
  Let $A$ and $A'$ be value objects, and let $c, c' : A \rel A'$ be relations
  between the underlying predomains. We say that $c$ and $c'$ are
  \textbf{quasi-equivalent}, written $c \qordeq c'$, if there exist
  syntactic perturbations $\delta^l_1, \delta^l_2 \in M_A$ and $\delta^r_1,
  \delta^r_2 \in M_{A'}$ such that there is a square $i_A(\delta^l_1)
  \ltsq{c}{c'} i_{A'}(\delta^r_1)$ and a square $i_A(\delta^l_2) \ltsq{c'}{c}
  i_{A'}(\delta^r_2)$. Given computation objects $B$ and $B$, we make the
  analogous definition for relations $d, d' : B \rel B'$ between the underlying
  error domains.
\end{definition}

@@

We can now give the final definition of value and computation relations for our
model, which serve as the denotation of our type precision derivations:
%
\begin{definition}
A \textbf{value relation} between value objects $(A, M_A, i_A)$ and $(A',
M_{A'}, i_{A'})$ consists of a predomain relation $c : A \rel A'$ that has a
push-pull structure, is quasi-left-representable by a morphism $e_c : A \to A'$,
and is such that $\li c$ is quasi-right-representable by an error domain morphism
$p_c : \li A' \to \li A$.

@@

To model the term precision in a way that is oblivious to the stepping
behavior of terms, we need to weaken the notion of square to allow for the
morphisms on either side to be ``out of sync'' rather than in lock step.
In the below, $f' \bisimsq{A_i}{A_o} f$ is the natural extension of
bisimilarity to morphisms: given bisimilar inputs, the functions have
bisimilar outputs.
\begin{definition}
  Let $c_i : A_i \rel A_i'$, $c_o : A_o \rel A_o'$ $f : A_i \to A_o$
  and $g : A_i' \to A_o'$.  An \textbf{extensional square} $f
  \ltsqbisim{c_i}{c_o} g$ holds if there \emph{exist} $f'
  \bisimsq{A_i}{A_o} f$ and $g' \bisimsq{A_i'}{A_o'} g$ such that $f' \ltsq{c_i}{c_o} g'$.
\end{definition}

@@

The term semantics $\sem{M}$ is nearly identical to the term semantics in
Section~\ref{sec:concrete-term-model}. One difference is that the interpretation
of the function type $A_i \ra A_o$ in the term semantics of Section
\ref{sec:concrete-term-model} involves \emph{arbitrary functions}, while the
interpretation of the function type in the term semantics defined here involves
only those functions that are morphisms of predomains. Additionally, the
semantics of casts is slightly different. In the semantics of Section
\ref{sec:concrete-term-model}, the projection corresponding to the type
precision $cc'$ is simply the composition of the projections. In the new
semantics, the corresponding projection involves a perturbation.  
%
% EXTENDED VERSION
% This is a consequence of the construction of the composition of
% value/computation relations (the relevant projection is defined in the proof
% of Lemma \ref{lem:representation-comp-F-U} in the Appendix).
%
% CONFERENCE VERSION
This is a consequence of the construction of the composition of
value/computation relations (the relevant projection is defined in the appendix
of the extended version \cite{gtt-sgdt-extended-version}).
%
However, the two projections are extensionally equivalent since they are the
same up to a perturbation.

### 2025-01-07 paper-new/concrete-term-model.tex
check in paper changes as of camera-ready submission on 11/22/24

For our simple denotational semantics, the category $\calV$ is simply
the category $\Set$ of sets and functions. Then because GTLC is
call-by-value, all types will be interpreted as objects of this
category, i.e., sets. More interesting are the computation
types. These will be interpreted in a category of sets equipped with
algebraic\footnote{The $\theta$ structure is not algebraic in the
strictest sense since it does not have finite arity.} structure with
morphisms being functions that are homomorphic in this structure.
This algebraic structure needs to model the effects present in our
language, in this case these are errors as well as taking
computational steps (potentially diverging by taking computational
steps forever).
%
We call these algebraic structures simple error domains to distinguish
them from the error domains we define later to model graduality:
\begin{definition}[Simple Error Domains]
  A (simple) error domain $B$ consists of
  \begin{enumerate}
  \item A carrier set $UB$
  \item An element $\mho_{B} : UB$ representing error
  \item A function $\theta_B : \laterhs UB \to UB$ modeling a computational step
  \end{enumerate}
  A homomorphism of error domains $\phi : B \multimap B'$ is a
  function $\phi : UB \to UB'$ that preserves $\mho$ and $\theta$, i.e.,
  $\phi(\mho_B) = \mho_{B'}$ and $\phi(\theta_{B}(\tilde{x})) =
  \theta_{B'}(\later (\phi(\tilde{x})))$.  Simple error domains and
  homomorphisms assemble into a category $\ErrDom$ with a forgetful
  functor $U : \ErrDom \to \Set$ taking the underlying set and
  function.
\end{definition}
Here is our first point where we utilize guarded type theory: rather
than simply being a function $UB \to UB$, the ``think'' map $\theta$
takes an element \emph{later}. This makes a major difference, because
the structure of a think map combined with the guarded fixpoint
operator allows us to define recursive elements of $UB$ in that any
function $f : UB \to UB$ has a ``quasi-fixed point'' $\textrm{qfix}(f)
= \fix(f \circ \theta_B)$ satisfying the quasi-fixed point property:
\[ \qfix(f) = f(\delta_B(\qfix f)) \]
where $\delta_B = \theta_B \circ \nxt$ is a map we call the ``delay''
map which trivially delays an element now to be available later.  As
an example, we can define $\Omega_B = \qfix(\id) : UB$, the
``diverging element'' that ``thinks forever'' in that $\Omega_B =
\delta_B(\Omega_B)$. We call this a ``quasi'' fixed point because it is
\emph{nearly} a fixed point except for the presence of the delay map
$\delta_B$, which is irrelevant from an extensional point of view
where we would prefer to ignore differences in the number of steps
that computations take.

@@

Then to model effectful programs, we need to construct a left adjoint
$\li : \Set \to \ErrDom$ to $U$ which constructs the ``free error
domain'' on a set, which models an effectful CBV computation.
%% \subsection{The Lift + Error Monad}\label{sec:lift-monad}
We base the construction on the \emph{guarded
lift monad} described in \cite{mogelberg-paviotti2016}. Here, we augment the
guarded lift monad to accommodate the additional effect of failure.
\begin{definition}[Free error domain]
  For a set $A$, we define the (carrier of) \emph{free error domain} $U(\li A)$ as the unique solution to the guarded domain equation:
  \[ U(\li A) \cong A + 1\, + \laterhs U(\li A). \]
  For which we use the following notation for the three constructors:
  \begin{enumerate}
  \item $\eta \colon A \to U(\li A)$
  \item $\mho \colon U(\li A)$
  \item $\theta \colon \laterhs U(\li A) \to U(\li A)$
  \end{enumerate}
  Here $\mho, \theta$ provide the error domain structure for $\li A$,
  and this has the universal property of being free in that any
  function $f : U(\li A) \to UB$ uniquely extends to a homomorphism
  $f^\dagger : \li A \multimap B$ satisfying $Uf^\dagger \circ \eta =
  f$. In other words, the free error monad is left adjoint $\li \dashv
  U$ to the forgetful functor, with $\eta$ the unit of the adjunction.
  %
  We will write $\theta_t(\dots)$ to mean $\theta (\lambda t. \dots)$.
\end{definition}
%% %
%% Unless otherwise mentioned, all constructs involving $\later$ or $\fix$ are
%% understood to be with respect to a fixed clock $k$. So for the above, we really
%% have for each clock $k$ a type $\li^k A$ with respect to that clock.
%% %
%% Formally, the lift monad $\li A$ is defined as the solution to the guarded
%% recursive type equation
%% %
%
An element of $\li A$ should be viewed as a computation that can either (1)
return a value (via $\eta$), (2) raise an error and stop (via $\mho$), or (3)
take a computational step (via $\theta$).
%
The operation $f^\dagger$ can be defined by guarded recursion in the
apparent way, and proven to satisfy the freeness property by L\"ob
induction.
% It is instructive to give at least one example of a use of guarded recursion, so
% we show below how to define extend:
% TODO
%
%
%% Verifying that the monadic laws hold uses \lob-induction and is straightforward.

>>>>

For our simple denotational semantics, the category $\calV$ is simply
the category $\Set$ of sets and functions. Then because GTLC is
call-by-value, all types will be interpreted as objects of this
category, i.e., sets. More interesting are the computation
types. These will be interpreted in a category of sets equipped with
algebraic\footnote{The $\theta$ structure is not algebraic in the
strictest sense since it does not have finite arity.} structure with
morphisms being functions that are homomorphic in this structure.
This algebraic structure needs to model the effects present in our
language, in this case these are errors as well as taking
computational steps (potentially diverging by taking computational
steps forever).
%
We call these algebraic structures simple error domains to distinguish
them from the error domains we define later to model graduality:
\begin{definition}
  A (simple) error domain $B$ consists of
  \begin{enumerate}
  \item A carrier set $UB$
  \item An element $\mho_{B} : UB$ representing error
  \item A function $\theta_B : \laterhs UB \to UB$ modeling a computational step
  \end{enumerate}
  A homomorphism of error domains $\phi : B \multimap B'$ is a
  function $\phi : UB \to UB'$ that preserves $\mho$ and $\theta$, i.e.,
  $\phi(\mho_B) = \mho_{B'}$ and $\phi(\theta_{B}(\tilde{x})) =
  \theta_{B'}(\later (\phi(\tilde{x})))$.  Simple error domains and
  homomorphisms assemble into a category $\ErrDom$ with a forgetful
  functor $U : \ErrDom \to \Set$ taking the underlying set and
  function.
\end{definition}
Here is our first point where we utilize guarded type theory: rather
than simply being a function $UB \to UB$, the ``think'' map $\theta$
takes an element \emph{later}. This makes a major difference, because
the structure of a think map combined with the guarded fixpoint
operator allows us to define recursive elements of $UB$ in that any
function $f : UB \to UB$ has a ``quasi-fixed point'' $\textrm{qfix}(f)
= \fix(f \circ \theta_B)$ satisfying the quasi-fixed point property:
\[ \qfix(f) = f(\delta_B(\qfix f)) \]
where $\delta_B = \theta_B \circ \nxt$ is a map we call the ``delay''
map which trivially delays an element now to be available later.  As
an example, we can define $\Omega_B = \qfix(\id) : UB$, the
``diverging element'' that ``thinks forever'' in that $\Omega_B =
\delta_B(\Omega_B)$. We call this a ``quasi'' fixed point because it is
\emph{nearly} a fixed point except for the presence of the delay map
$\delta_B$, which is irrelevant from an extensional point of view
where we would prefer to ignore differences in the number of steps
that computations take.

@@

Then to model effectful programs, we need to construct a left adjoint
$\li : \Set \to \ErrDom$ to $U$ which constructs the ``free error
domain'' on a set, which models an effectful CBV computation.
%% \subsection{The Lift + Error Monad}\label{sec:lift-monad}
We base the construction on the \emph{guarded
lift monad} described in \cite{mogelberg-paviotti2016}. Here, we augment the
guarded lift monad to accommodate the additional effect of failure.
\begin{definition}
  For a set $A$, we define the (carrier of) \emph{free error domain} $U(\li A)$ as the unique solution to the guarded domain equation:
  \[ U(\li A) \cong A + 1\, + \laterhs U(\li A). \]
  For which we use the following notation for the three constructors:
  \begin{enumerate}
  \item $\eta \colon A \to U(\li A)$
  \item $\mho \colon U(\li A)$
  \item $\theta \colon \laterhs U(\li A) \to U(\li A)$
  \end{enumerate}
  Here $\mho, \theta$ provide the error domain structure for $\li A$,
  and this has the universal property of being free in that any
  function $f : U(\li A) \to UB$ uniquely extends to a homomorphism
  $f^\dagger : \li A \multimap B$ satisfying $Uf^\dagger \circ \eta =
  f$. In other words, the free error monad is left adjoint $\li \dashv
  U$ to the forgetful functor, with $\eta$ the unit of the adjunction.
  %
  We will write $\theta_t(\dots)$ to mean $\theta (\lambda t. \dots)$.
\end{definition}
%% %
%% Unless otherwise mentioned, all constructs involving $\later$ or $\fix$ are
%% understood to be with respect to a fixed clock $k$. So for the above, we really
%% have for each clock $k$ a type $\li^k A$ with respect to that clock.
%% %
%% Formally, the lift monad $\li A$ is defined as the solution to the guarded
%% recursive type equation
%% %
%
An element of $\li A$ should be viewed as a computation that can either (1)
return a value (via $\eta$), (2) raise an error and stop (via $\mho$), or (3)
take a computational step (via $\theta$).
%
The operation $f^\dagger$ can be defined by guarded recursion in the
apparent way, and proven to satisfy the freeness property by L\"ob
induction.
% It is instructive to give at least one example of a use of guarded recursion, so
% we show below how to define extend:
% TODO
%
%
%% Verifying that the monadic laws hold uses \lob-induction and is straightforward.

### 2025-01-07 paper-new/discussion.tex
cite our artifact's zenodo doi
check in paper changes as of camera-ready submission on 11/22/24

\subsection{Mechanization}\label{sec:mechanization}
% Discuss Guarded Cubical Agda and mechanization efforts
In parallel with developing the theory discussed in this paper, we have
developed a partial formalization of our results in Guarded Cubical Agda
\cite{veltri-vezzosi2020}.
%
We formalized the major components of the definition of the concrete model
described in Section \ref{sec:concrete-relational-model}: predomains, error
domains, morphisms, relations, squares, using the guarded
features to define the free error domain, the lock-step error ordering, and weak
bisimilarity. We also formalized the no-go theorem from Section
\ref{sec:towards-relational-model}.

@@

The remaining formalization work is the following:
\begin{enumerate}
    \item Showing that the functors $\times$ and $\arr$ preserve
    quasi-representability, which requires tedious reasoning about the Kleisli actions of
    $\times$ and $\arr$. These results are proved in the Appendix.
    \item Verifying the rules in the model corresponding to the equations for
    type precision derivations (see the bottom of Figure \ref{fig:gtlc-syntax}).
    These are also included in the Appendix.
    \item Formalizing the syntax-to-semantics translation.
\end{enumerate}

@@

This work has focused on the ``Natural'' semantics of casts , which
validate the full call-by-value $\eta$ laws for types, but other cast
semantics have been proposed such as eager
(\cite{herman-tomb-flanagan-2010}) and transient (\cite{transient})
cast semantics which trade off certain $\eta$ equalities for other
benefits such as early error detection or reduced runtime
overhead\cite{deepandshallowtypes,new-licata-ahmed2019}. It would be a
good test of the generality of our semantic framework to see if it can
model these alternative semantics by adjusting the semantics of types
and casts, and providing proofs of weakened $\eta$ principles.

>>>>

\subsection{Mechanization}\label{sec:mechanization}
% Discuss Guarded Cubical Agda and mechanization efforts
In parallel with developing the theory discussed in this paper, we have
developed a partial formalization of our results in Guarded Cubical Agda
\cite{veltri-vezzosi2020}; the formalization is available online \cite{artifact}.
%
We formalized the major components of the definition of the concrete model
described in Section \ref{sec:concrete-relational-model}: predomains, error
domains, morphisms, relations, squares, using the guarded
features to define the free error domain, the lock-step error ordering, and weak
bisimilarity. We also formalized the no-go theorem from Section
\ref{sec:towards-relational-model}.

@@

As of this writing, the remaining formalization work is the following:
\begin{enumerate}
    \item Showing that the functors $\times$ and $\arr$ preserve
    quasi-representability, which requires tedious reasoning about the Kleisli actions of
    $\times$ and $\arr$. These results are proved in the Appendix.
    \item Verifying the rules in the model corresponding to the equations for
    type precision derivations (see the bottom of Figure \ref{fig:gtlc-syntax}).
    These are also included in the Appendix.
    \item Formalizing the syntax-to-semantics translation.
\end{enumerate}

@@

This work has focused on the ``Natural'' semantics of casts , which
validate the full call-by-value $\eta$ laws for types, but other cast
semantics have been proposed such as eager
(\cite{herman-tomb-flanagan-2010}) and transient (\cite{transient})
cast semantics which trade off certain $\eta$ equalities for other
benefits such as early error detection or reduced runtime
overhead \cite{deepandshallowtypes,new-licata-ahmed2019}. It would be a
good test of the generality of our semantic framework to see if it can
model these alternative semantics by adjusting the semantics of types
and casts, and providing proofs of weakened $\eta$ principles.

### 2025-01-07 paper-new/intro.tex
cite our artifact's zenodo doi
check in paper changes as of camera-ready submission on 11/22/24

Providing a semantics for gradual typing is inherently complicated in
that it involves: (1) recursion and recursive types through the
presence of dynamic types, (2) effects in the form of divergence and
errors (3) relational models in capturing the graduality
property. Recursion and recursive types must be handled using some
flavor of domain theory. Effects can be modeled using monads in the
style of Moggi, or adjunctions in the style of
Levy\cite{MOGGI199155,39155,levy99,levystacks}. Relational properties and their verification
lead naturally to the use of reflexive graph categories or double
categories\cite{dunphythesis,new-licata18}.

@@

The main contribution of this work is a compositional denotational semantics in
guarded type theory for a simple gradually typed language that validates
$\beta\eta$ equality and satisfies a graduality theorem.
%
% In particular, the notion of syntactic perturbations stands out as a key
% technical contribution and is the most significant and novel feature of our
% model that allowed us to successfully adapt the denotational approach of New and
% Licata to the guarded setting.
% In particular, the notion of syntactic perturbations stands out as a key feature
% of our model, and is our most significant technical contribution allowing us to
% successfully adapt the denotational approach of New and Licata to the guarded
% setting.
%
Within our semantics, the notion of syntactic perturbation is our most
significant technical contribution, allowing us to successfully adapt the
denotational approach of New and Licata to the guarded setting.
%
Most of our work has further been verified in Guarded Cubical Agda
\cite{veltri-vezzosi2020}, demonstrating that the semantics is readily
mechanizable. We provide an overview of the mechanization effort, and what
remains to be formalized, in Section \ref{sec:mechanization}.
%

>>>>

Providing a semantics for gradual typing is inherently complicated in
that it involves: (1) recursion and recursive types through the
presence of dynamic types, (2) effects in the form of divergence and
errors (3) relational models in capturing the graduality
property. Recursion and recursive types must be handled using some
flavor of domain theory. Effects can be modeled using monads in the
style of Moggi, or adjunctions in the style of
Levy \cite{MOGGI199155,39155,levy99,levystacks}. Relational properties and their verification
lead naturally to the use of reflexive graph categories or double
categories \cite{dunphythesis,new-licata18}.

@@

The main contribution of this work is a compositional denotational semantics in
guarded type theory for a simple gradually typed language that validates
$\beta\eta$ equality and satisfies a graduality theorem.
%
% In particular, the notion of syntactic perturbations stands out as a key
% technical contribution and is the most significant and novel feature of our
% model that allowed us to successfully adapt the denotational approach of New and
% Licata to the guarded setting.
% In particular, the notion of syntactic perturbations stands out as a key feature
% of our model, and is our most significant technical contribution allowing us to
% successfully adapt the denotational approach of New and Licata to the guarded
% setting.
%
Within our semantics, the notion of syntactic perturbation is our most
significant technical contribution, allowing us to successfully adapt the
denotational approach of New and Licata to the guarded setting.
%
Most of our work has further been verified in Guarded Cubical Agda
\cite{veltri-vezzosi2020}, demonstrating that the semantics is readily
mechanizable. The Agda formalization is available online \cite{artifact}.
We provide an overview of the mechanization effort, and what
remains to be formalized, in Section \ref{sec:mechanization}.
%

### 2025-01-07 paper-new/paper.tex
remove two of the formatting commands from paper.tex
check in paper changes as of camera-ready submission on 11/22/24

\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newenvironment{DIFnomarkup}{}{}
\pagestyle{empty}
\settopmatter{printfolios=false}

%% CCS Concepts
\begin{CCSXML}
  <ccs2012>
     <concept>
         <concept_id>10003752.10010124.10010131.10010133</concept_id>
         <concept_desc>Theory of computation~Denotational semantics</concept_desc>
         <concept_significance>500</concept_significance>
         </concept>
     <concept>
         <concept_id>10011007.10011006.10011039.10011311</concept_id>
         <concept_desc>Software and its engineering~Semantics</concept_desc>
         <concept_significance>300</concept_significance>
         </concept>
     <concept>
         <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
         <concept_desc>Software and its engineering~Functional languages</concept_desc>
         <concept_significance>300</concept_significance>
         </concept>
   </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Denotational semantics}
\ccsdesc[300]{Software and its engineering~Semantics}
\ccsdesc[300]{Software and its engineering~Functional languages}

@@

\newpage
\appendix
\input{appendix}

>>>>

% \theoremstyle{remark} 
% \newtheorem{theorem}{Theorem}
% \newtheorem{corollary}{Corollary}
\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newenvironment{DIFnomarkup}{}{}
% \pagestyle{empty}
% \settopmatter{printfolios=false}

%% CCS Concepts
\begin{CCSXML}
  <ccs2012>
     <concept>
         <concept_id>10003752.10010124.10010131.10010133</concept_id>
         <concept_desc>Theory of computation~Denotational semantics</concept_desc>
         <concept_significance>500</concept_significance>
         </concept>
     <concept>
         <concept_id>10011007.10011006.10011039.10011311</concept_id>
         <concept_desc>Software and its engineering~Semantics</concept_desc>
         <concept_significance>500</concept_significance>
         </concept>
     <concept>
         <concept_id>10011007.10011006.10011008.10011009.10011012</concept_id>
         <concept_desc>Software and its engineering~Functional languages</concept_desc>
         <concept_significance>300</concept_significance>
         </concept>
   </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Denotational semantics}
\ccsdesc[500]{Software and its engineering~Semantics}
\ccsdesc[300]{Software and its engineering~Functional languages}

@@

% \newpage
% \appendix
% \input{appendix}
